---
title: Mini Project01
format: html
css: styles.css
---

# Introduction
This project for STA 9750 has us looking at Federal Transit Administration Data, specifically for 2022.

We will try to answer some questions about transit from that year in this file.

# Import Libraries

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
```

# Import Data
```{r echo=TRUE, message=FALSE, warning=FALSE}
# I had to rename several of the files on import,
# When re-running my code, please be mindful of that
FARES <- readxl::read_xlsx("2022_fare_revenue.xlsx") |>
  select(-`State/Parent NTD ID`, 
         -`Reporter Type`,
         -`Reporting Module`,
         -`TOS`,
         -`Passenger Paid Fares`,
         -`Organization Paid Fares`) |>
  filter(`Expense Type` == "Funds Earned During Period") |>
  select(-`Expense Type`)

EXPENSES <- readr::read_csv("2022_expenses.csv") |>
  select(`NTD ID`, 
         `Agency`,
         `Total`, 
         `Mode`) |>
  mutate(`NTD ID` = as.integer(`NTD ID`)) |>
  rename(Expenses = Total) |>
  group_by(`NTD ID`, `Mode`) |>
  summarize(Expenses = sum(Expenses)) |>
  ungroup()

FINANCIALS <- inner_join(FARES, EXPENSES, join_by(`NTD ID`, `Mode`))

TRIPS <- readxl::read_xlsx("ridership.xlsx", sheet="UPT") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="UPT") |>
  drop_na() |>
  mutate(month=my(month)) # Parse _m_onth _y_ear date specs

MILES <- readxl::read_xlsx("ridership.xlsx", sheet="VRM") |>
  filter(`Mode/Type of Service Status` == "Active") |>
  select(-`Legacy NTD ID`, 
         -`Reporter Type`, 
         -`Mode/Type of Service Status`, 
         -`UACE CD`, 
         -`TOS`) |>
  pivot_longer(-c(`NTD ID`:`3 Mode`), 
               names_to="month", 
               values_to="VRM") |>
  drop_na() |>
  group_by(`NTD ID`, `Agency`, `UZA Name`, 
           `Mode`, `3 Mode`, month) |>
  summarize(VRM = sum(VRM)) |>
  ungroup() |>
  mutate(month=my(month)) # Parse _m_onth _y_ear date specs

USAGE <- inner_join(TRIPS, MILES) |>
  mutate(`NTD ID` = as.integer(`NTD ID`))
```


```{r echo=TRUE, message=FALSE}
if(!require("DT")) install.packages("DT")
library(DT)
sample_n(USAGE, 1000) |> 
  mutate(month=as.character(month)) |> 
  DT::datatable()
```

## Task 1: Rename column UZA Name to metro_area
```{r echo=TRUE, message=FALSE}
colnames(USAGE)

USAGE <- USAGE |> rename("metro_area" = "UZA Name")

colnames(USAGE)
```

## Task 2: Find unique Modes, impute with a more understandable value using case-when

```{r message=TRUE}
head(USAGE)
USAGE |>
  distinct(Mode) |>
  arrange(Mode)


USAGE <- USAGE |>
  mutate(Mode=case_when(
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "CC" ~ "Cable Car",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferryboat",
    Mode == "HR" ~ "Heavy Rail",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "LR" ~ "Light Rail",
    Mode == "MB" ~ "Bus",
    Mode == "MG" ~ "Monorail and Automated Guideway modes",
    Mode == "PB" ~ "Publico",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "TR" ~ "Aerial Tramways",
    Mode == "VP" ~ "Vanpool",
    Mode == "YR" ~ "Hybrid Rail",
    TRUE ~ "Unknown"))

head(USAGE)
USAGE |>
  distinct(Mode) |>
  arrange(Mode)

USAGE <- USAGE |>     rename( "unlinked_passenger_trips" = "UPT",
                              "vehicle_revenue_miles" = "VRM") |>
  select( -`NTD ID`,
          -`3 Mode`)

sample_n(USAGE, 1000) |> 
    mutate(month=as.character(month)) |> 
    DT::datatable()
```

## Task 3: Answer specific questions
``` {r message=TRUE}
head(USAGE)
```

### 3.1: Which transit agency had the most total VRM in this sample?
``` {r echo=TRUE, message=TRUE}
# A grouped summarize that takes the first returned item in descending order should get that
USAGE |>
  group_by(Agency) |>
  summarize(n_vrm = n()) |>
  arrange(desc(n_vrm)) |>
  slice_head(n=1)
```

Answer: New Jersey Transit Corporation with 2,048

### 3.2: Which transit mode had the most total VRM in this sample?
``` {r echo=TRUE, message=TRUE}
# Same code as 3.1 but replacing Agency with Mode
USAGE |>
  group_by(Mode) |>
  summarize(n_vrm = n()) |>
  arrange(desc(n_vrm)) |>
  slice_head(n=1)
```

Answer: Demand Response with 115,701

### 3.3: How many trips were taken on the NYC Subway (Heavy Rail) in May 2024?

``` {r echo=TRUE, message=TRUE}
#install.packages("stringr")
library(stringr)

# First checking to ensure that May 2024 is only denoted by 2024-05-01
USAGE |>
  filter(str_detect(month, "2024-05-")) |>
  distinct(month) #checks out

# Ensuring I get the right Agency name
USAGE |>
  filter(str_detect(Agency,"MTA")) |>
  distinct(Agency)

# Unlinked Passenger Trips (UPT) are defined as:
# The number of passengers who board public transportation vehicles.
# Passengers are counted each time they board vehicles
# no matter how many vehicles they use to travel from their origin
# to their destination.


USAGE |>
  filter(month == '2024-05-01',
         Agency == "MTA New York City Transit",
         Mode == "Heavy Rail") |>
  select(unlinked_passenger_trips)
```

Answer: 180,458,819 trips


### 3.4: Which Mode of transport had the longest average trip in May 2024?

``` {r echo=TRUE, message=FALSE}
USAGE |>
  filter(month == '2024-05-01') |>
  group_by(Mode) |>
  summarize(trips_avg = mean(unlinked_passenger_trips)) |>
  arrange(desc(trips_avg)) |>
  slice_head(n=1)
```
 Answer: Heavy Rail, with 14,836,486 average trips
(I did this question before 3.4 was marked as unneccesary due to not having the correct feature to run this analysis, but I've left it in regardless)

### 3.5: How much did NYC Subway ridership fall between April 2019 and April 2020?

``` {r echo=TRUE, message=TRUE}
USAGE |>
  filter(Agency == "MTA New York City Transit",
         Mode == "Heavy Rail",
         month >= '2019-04-01',
         month < '2020-05-01') |>
  select(unlinked_passenger_trips)

# Let's plot the ridership numbers
library(ggplot2)
library(lubridate)

dates <- seq(ymd("2019-04-01"), by = "month", length.out = 13)
ridership <- c(USAGE |>
                 filter(month >= '2019-04-01',
                        month < '2020-05-01',
                        Agency == 'MTA New York City Transit',
                        Mode == 'Heavy Rail'))

# Create a data frame
data <- data.frame(Date = dates, Ridership = ridership)

# install.packages("scales")
library(scales)

# Plot ridership over time
ggplot(data, aes(x = Date, y = Ridership.unlinked_passenger_trips)) +  # Change x and y accordingly
  geom_line() +
  geom_point() +
  labs(title = "NYC Subway Ridership from April 2019 through April 2020", x = "Date", y = "Total Trips") +
  scale_y_continuous(labels = label_number()) +  # Prevent scientific notation
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  # Center the title and make it bold
    axis.title.x = element_text(size = 14),  # Change x-axis title size
    axis.title.y = element_text(size = 14),  # Change y-axis title size
    axis.text = element_text(size = 12),  # Change axis text size
    panel.grid.minor = element_blank()  # Remove minor grid lines
  )
```

## Task 4: Find 3 more interesting facts in this data
``` {r echo=TRUE, message=TRUE}
head(USAGE, n=7)
```

### 4.1: What's the most miles traveled by agency + mode and how does it compare to the number of trips?
``` {r echo=TRUE, message=TRUE}
USAGE |>
  group_by(Agency, Mode) |>
  summarize(total_travel = sum(vehicle_revenue_miles, na.rm = TRUE), .groups="drop") |>
  arrange(desc(total_travel)) |>
  slice_head(n=3)
```
NYC Subway leads the way in revenue miles, followed by NJT Bus and LA Bus

``` {r echo=TRUE, message=TRUE}
USAGE |>
  group_by(Agency, Mode) |>
  summarize(total_travel = sum(unlinked_passenger_trips, na.rm = TRUE), .groups="drop") |>
  arrange(desc(total_travel)) |>
  slice_head(n=3)
```
NYC Subway leads the way here as well, while NJT Bus is overtaken by NYC's bus system. The distance between the MTA and the next few is quite large.


### 4.2 Which Agency manages the most Modes?
``` {r echo=TRUE, message=TRUE}
USAGE |>
  group_by(Agency) |>
  summarize(total_modes = n_distinct(Mode)) |>
  arrange(desc(total_modes)) # Massachusetts Bay Transportation Authority
```

### 4.3 Which date saw the most usage for the NYC Subway?
``` {r echo=TRUE, message=TRUE}
USAGE |>
  filter(Agency == 'MTA New York City Transit',
         Mode == 'Heavy Rail') |>
  group_by(month) |>
  summarize(most_trips = sum(unlinked_passenger_trips, na.rm = TRUE)) |>
  arrange(desc(most_trips)) |>
  slice_head(n=1) # 2019-10-01

# conversely, which had the fewest?
USAGE |>
  filter(Agency == 'MTA New York City Transit',
         Mode == 'Heavy Rail') |>
  group_by(month) |>
  summarize(most_trips = sum(unlinked_passenger_trips, na.rm = TRUE)) |>
  arrange((most_trips)) |>
  slice_head(n=1) # 2020-04-01
```
2019-10-01 had the most while 2020-04-01 had the fewest

## Task 5: Create a new table with annual total (sum) for UPT and VRM for 2022
``` {r echo=TRUE, message=TRUE}
USAGE <- inner_join(TRIPS, MILES) |>
  mutate(`NTD ID` = as.integer(`NTD ID`))


USAGE <- USAGE |>
  mutate(Mode=case_when(
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "CC" ~ "Cable Car",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferryboat",
    Mode == "HR" ~ "Heavy Rail",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "LR" ~ "Light Rail",
    Mode == "MB" ~ "Bus",
    Mode == "MG" ~ "Monorail and Automated Guideway modes",
    Mode == "PB" ~ "Publico",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "TR" ~ "Aerial Tramways",
    Mode == "VP" ~ "Vanpool",
    Mode == "YR" ~ "Hybrid Rail",
    TRUE ~ "Unknown"))


USAGE_2022_ANNUAL <- USAGE |>
  select(-`3 Mode`) |>
  rename("metro_area" = "UZA Name") |>
  filter(year(month)=='2022') |>
  group_by(`NTD ID`, `Agency`,`metro_area`,`Mode`) |>
  summarize(UPT = sum(UPT),
            VRM = sum(VRM)) |>
  ungroup()

head(USAGE_2022_ANNUAL, n=2)

FINANCIALS <- FINANCIALS |>
  mutate(Mode=case_when(
    Mode == "AR" ~ "Alaska Railroad",
    Mode == "CB" ~ "Commuter Bus",
    Mode == "CC" ~ "Cable Car",
    Mode == "CR" ~ "Commuter Rail",
    Mode == "DR" ~ "Demand Response",
    Mode == "FB" ~ "Ferryboat",
    Mode == "HR" ~ "Heavy Rail",
    Mode == "IP" ~ "Inclined Plane",
    Mode == "LR" ~ "Light Rail",
    Mode == "MB" ~ "Bus",
    Mode == "MG" ~ "Monorail and Automated Guideway modes",
    Mode == "PB" ~ "Publico",
    Mode == "RB" ~ "Bus Rapid Transit",
    Mode == "SR" ~ "Streetcar Rail",
    Mode == "TB" ~ "Trolleybus",
    Mode == "TR" ~ "Aerial Tramways",
    Mode == "VP" ~ "Vanpool",
    Mode == "YR" ~ "Hybrid Rail",
    TRUE ~ "Unknown"))


USAGE_AND_FINANCIALS <- left_join(USAGE_2022_ANNUAL,
                                  FINANCIALS,
                                  join_by(`NTD ID`, `Mode`)) |>
  drop_na()


head(USAGE_AND_FINANCIALS,n=7)
```

## Task 6: Answer the following 6 questions using USAGE_AND_FINANCIALS

### 6.1: Which transit system (agency and mode) had the most UPT in 2022?

``` {r echo=TRUE, message=TRUE}
USAGE_AND_FINANCIALS |>
  group_by(`Agency`, `Mode`) |>
  summarize(Total_UPT = sum(UPT, na.rm=TRUE), .groups="drop") |>
  arrange(desc(Total_UPT)) |>
  slice_head(n=1)
```
MTA New York City Transit -- Heavy Rail


### 6.2: Which transit system (agency and mode) had the highest farebox recovery, defined as the highest ratio of Total Fares to Expenses?
``` {r echo=TRUE, message=TRUE}
USAGE_AND_FINANCIALS |>
  group_by(`Agency`,`Mode`) |>
  summarize(fbx_rcv = sum(`Total Fares`, na.rm = TRUE) / sum(`Expenses`, na.rm = TRUE), .groups="drop") |>
  arrange(desc(fbx_rcv)) |>
  slice_head(n=1)
```
Transit Authority of Central Kentucky -- Vanpool

### 6.3: Which transit system (agency and mode) had the lowest expenses per UPT?
``` {r echo=TRUE, message=TRUE}
USAGE_AND_FINANCIALS |>
  filter(UPT >= 1) |>
  group_by(`Agency`,`Mode`) |>
  summarize(lwst_exp_upt =  sum(`Expenses`, na.rm = TRUE) / sum(`UPT`, na.rm = TRUE), .groups="drop") |>
  arrange(lwst_exp_upt) |>
  slice_head(n=1)
```
North Caroline State University -- Bus
  

### 6.4: Which transit system (agency and mode) had the highest total fares per UPT?
``` {r echo=TRUE, message=TRUE}
USAGE_AND_FINANCIALS |>
  filter(`Total Fares` >= 1) |>
  group_by(`Agency`,`Mode`) |>
  summarize(hghst_fr_upt = sum(`UPT`, na.rm = TRUE) / sum(`Total Fares`, na.rm = TRUE), .groups="drop") |>
  # summarize(hghst_fr_upt = sum(`UPT`, na.rm = TRUE) / sum(`Total Fares`, na.rm = TRUE), .groups="drop")  |>
  arrange(desc(hghst_fr_upt)) |>
  slice_head(n=1)
```
Athens-Clarke County Unified Government -- Bus
  

### 6.5: Which transit system (agency and mode) had the lowest expenses per VRM?
``` {r echo=TRUE, message=TRUE}
USAGE_AND_FINANCIALS |>
  filter(UPT >= 1) |>
  group_by(`Agency`,`Mode`) |>
  summarize(lwst_exp_vrm = sum(`Expenses`, na.rm = TRUE) / sum(`VRM`, na.rm = TRUE), .groups="drop") |>
  arrange(lwst_exp_vrm) |>
  slice_head(n=1)
```
New Mexico Department of Transportation -- Vanpool

### 6.6: Which transit system (agency and mode) had the highest total fares per VRM?
``` {r echo=TRUE, message=TRUE}
USAGE_AND_FINANCIALS |>
  filter(`Total Fares` >= 1) |>
  group_by(`Agency`,`Mode`) |>
  summarize(hghst_fr_vrm = sum(`VRM`, na.rm = TRUE) / sum(`Total Fares`, na.rm = TRUE), .groups="drop") |>
  arrange(desc(hghst_fr_vrm)) |>
  slice_head(n=1)
```
Athens-Clarke County Unified Government -- Bus